---
title: "Assignment 5"
format: pdf
author:
  - name: "Kerem Karagöz"
  - name: "Immanuel Klein"
execute: 
  cache: true
editor_options: 
  chunk_output_type: console
editor: 
  markdown: 
    wrap: sentence
---

# General Information

-   **Points**: Assignment 5 comprises of 6 tasks, 2 points each (12 in total).
    2 points are obtained for complete and correct answers.
    1 point is obtained for a proper approach or if only part of the task is solved.

-   **Submission**: Hand in the assignment as a `Markdown` report ([RMarkdown](https://rmarkdown.rstudio.com/lesson-15.html) or [Quarto](https://rstudio.github.io/cheatsheets/html/quarto.html)) rendered as PDF.
    The PDF report should show the result(s), the code that produced the result(s), and possibly additional text or comment.
    Also indicate your name.
    The report should be uploaded on Moodle until Friday, July 14, 6 pm.

-   **Working in teams**: Everyone needs to hand in a report on Moodle.
    However, the report can be handed in as a team work (max. 2 people).
    When working in teams, state at the beginning of the document, who you worked with.
    It Ideally, teams use GitHub and add a link to the GitHub repository to which both contributed.

\newpage

# Additional remarks

## `ulam` and `Quarto`

Running MCMC with `ulam()` takes some time and produces many messages.
This can be annoying when you repeatedly render the `Quarto` document and the goal is to have a clean report.
Here are some tips to avoid long rendering times and an ugly document:

-   Write the ulam() model in a separate code chunk and give the chunk a name `{r name}`.In addition, specify the following settings at the top of the code chunk to avoid printing the MCMC progress messages of `ulam()` in the PDF document,

``` r
#| echo: true
#| eval: true
#| output: false

# data list and model
```

-   Set the [caching option](https://quarto.org/docs/computations/caching.html) `cache: true` in the YAML header at the start of the document. The first time the code chunk is rendered, its results are cached (stored) in the background. If you leave the code chunk untouched, the results are directly retrieved the next time you render the document. This avoids rerunning a model with every new rendering of the document. The chunk will only be newly evaluated if you actually change its code. To allow `Quarto` to recognize a code chunk, it needs a name.

## Data list and index variables

-   Provide `ulam()` a list of variables and values that you need for estimating the model rather than the entire data frame.
    `ulam()` works more reliable with lists.
    Moreover, while creating the list, you can also recode variables (e.g., from dummy values (0,1) or group names (male, female) to index values (1,2))

-   Using indices for group specific parameters such as in `y = a[G] + b[G]*X`, only works for integer values larger than 0, that is: `G = {1,2,3,...}`.
    When the variable values are names or include 0, you have to recode it first.
    Use the function `as.integer(variable)` if the variable is of type `factor` or `as.integer(as.factor(variable))` if the variable is of type `character`.
    Use `variable + 1` when the variable is dummy coded with 0 and 1 to get values 1 and 2.

\newpage

GitHub: <https://github.com/immanuel-klein/bayesian-assignments.git>

```{r packages}
# load packages here
library(dplyr)
library(tidyverse)
library(ggplot2)
library(tinytex)
library(rethinking)
library(rstan)
library(reshape2)
library(ggridges)
```

# Task Set 1

Load the data set `RiskyChoice.csv` to solve the Task Set 1.
Use the `read_csv2()` function instead of `read_csv()`.

```{r data}
# load data here
#Used read.csv2 instead of read_csv2 to correctly get the decimal seperators for NegativeAffect
#Then use as numerics to convert to num from character
risky <- read.csv2("RiskyChoice.csv")
risky$NegativeAffect <- as.numeric(risky$NegativeAffect)
head(risky)
```

## Task 1.1

Create a reduced data table with only one row per subject that shows the number of solved choices problems (`nChoice`) and the number of correct choices (`nCorrect`) for each subject along with the other variables.
Remove the subjects with missing values.
Print the data of the first 10 subjects.

```{r}
# write code here
risky.reduced <- risky %>%
  filter(!is.na(CorrectChoice)) %>%
  group_by(Subject) %>%
  summarise(
    Gender = first(Gender),
    NegativeAffect = first(NegativeAffect),
    Numeracy = first(Numeracy),
    nChoice = n(),
    nCorrect = sum(CorrectChoice)
  ) %>% na.omit()

head(risky.reduced, 10)
```

**Remark:** We understood the task the following way: If a subject has at least one NA value for CorrectChoice, do not take any values of this subject into the new table.
However, this left us with a tibble with the dimensions 0x3, i.
e.
there is no subject that has no NA value for CorrectChoice.
Thus, we decided to just filter the rows with NA for CorrectChoice instead of the whole subject.
Also, because risky.reduced should only have one row per subject, we could only include Gender, NegativeAffect, and Numeracy.
The others have differing values within the subjects.
We then omit NAs.

## Task 1.2

Run a Bayesian regression model that predicts `nCorrect` from `Numeracy` using fixed intercepts and fixed slopes.
Standardize the predictor before running the model and compute the WAIC of the model.

```{r m1}
#| echo: true
#| eval: true
#| output: false

# write data list and model here
risky.reduced.regr <- risky.reduced %>%
  mutate(
    Numeracy_std = (Numeracy - mean(Numeracy)) / sd(Numeracy),
    Subject_idx = as.numeric(as.factor(Subject))  # Convert Subject to a numeric index
)

data.list <- list(
  nCorrect = risky.reduced.regr$nCorrect,
  nChoice = risky.reduced.regr$nChoice,
  Numeracy_std = risky.reduced.regr$Numeracy_std
)

m1 <- ulam(
  alist(
    # Choosing intercept and slope:
    # When numeracy is 0, it's reasonable that correct choices come from guessing.
    # Thus, ca 50% of all choices might be correct (0.5 of ~ 100 choices).
    # Looking at the values where numeracy is 10, a 0.3 slope might make sense.
    nCorrect ~ dbinom(nChoice, p),
    logit(p) <- a + b * Numeracy_std,  # Logit link function
    a ~ dnorm(50, 10),                 # Prior for intercept on logit scale
    b ~ dnorm(0.3, 0.1)                # Prior for slope on logit scale
  ), data = data.list, chains = 4, cores = 4, log_lik = TRUE
)

# Summarize the model
precis(m1, depth = 2)
```

```{r}
#write code here
WAIC(m1)
```

## Task 1.3

Run a Bayesian regression model that predicts `nCorrect` from `Numeracy` using random intercepts and fixed slopes.
Standardize the predictor before running the model and compute the WAIC of the model.

```{r m2}
#| echo: true
#| eval: true
#| output: false

# write data list and model here

data.list2 <- list(
  nCorrect = risky.reduced.regr$nCorrect,
  nChoice = risky.reduced.regr$nChoice,
  Numeracy_std = risky.reduced.regr$Numeracy_std,
  Subject = risky.reduced.regr$Subject_idx
)

m2 <- ulam(
  alist(
    nCorrect ~ dbinom(nChoice, p),
    logit(p) <- a[Subject] + beta * Numeracy_std,  # Specifies the linear predictor with random intercepts a[Subject] for each subject and a fixed slope beta
    a[Subject] ~ dnorm(50, 10),  # Normal prior for the random intercepts a[Subject]
    beta ~ dnorm(0.3, 0.1)  # Normal prior for the slope beta
  ), data = data.list2, chains = 4, cores = 4, log_lik = TRUE
)

# Summarize the model
precis(m2, depth = 2)
```

```{r}
# write code here
WAIC(m2)
```

\newpage

# Task Set 2

## Task 2.1

Create a data table that entails 10,000 posterior samples (rows) for each subject-specific (columns) intercept.
Convert the sampled values into probabilities and print the first 10 samples of the first 10 subjects.

```{r}
# write code here
samples <- extract.samples(m2, n = 10000) 

# Convert intercepts to probabilities
prob.samples <- inv_logit(samples$a)

# Convert to data frame and print the first 10 samples for the first 10 subjects
prob.samples.df <- as.data.frame(prob.samples)
colnames(prob.samples.df) <- paste0("Subject.", 1:ncol(prob.samples.df))

head(prob.samples.df[, 1:10], 10)

```

## Task 2.2

Use the posterior samples to plot the posterior distribution of all subject-specific intercepts to show the variability in the performance among subjects.
Use the converted values (probabilities).

```{r}
# write code here
# As there are 119 different subjects it is quite hard to visualize the intercepts

# Convert the wide format to long format for ggplot2
prob.samples.long <- melt(prob.samples.df)

# Create the ridgeline plot
ggplot(prob.samples.long, aes(x = value, y = variable, height = ..density..)) +
  geom_density_ridges(stat = "density", scale = 2, alpha = 0.7) +
  labs(title = "Posterior Distributions of Subject-Specific Intercepts",
       x = "Probability", y = "Subject") +
  theme_ridges() +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())

# Plot the posterior distributions
ggplot(prob.samples.long, aes(x = value, fill = variable)) +
  geom_density(alpha = 0.5) +
  labs(title = "Posterior Distributions of Subject-Specific Intercepts",
       x = "Probability", y = "Density") +
  theme_minimal() +
  theme(legend.position = "none")

# An alternative version where we only use the first 10 subjects
prob.samples.subset <- prob.samples.df[, 1:10]
prob.samples.melt.subset <- melt(prob.samples.subset)

# Create a faceted plot for the first 10 subjects
ggplot(prob.samples.melt.subset, aes(x = value)) +
  geom_density(alpha = 0.5) +
  facet_wrap(~ variable, scales = "free_y") +
  labs(title = "Posterior Distributions of Subject-Specific Intercepts",
       x = "Probability", y = "Density") +
  theme_minimal()


```

## Task 2.3

Consider the following posterior summaries and traceplots.
Which model was estimated and what might be the cause of the convergence problems?

```{r}
#precis(m3)
#traceplot_ulam(m3, pars = c("mu_a", "tau_a", "mu_b", "tau_b"))

print("From the precis we can see that mu_a shows a good convergence with high effective sample size and Rhat close to 1, meaning that chains are well mixed. When we go from mu_a to tau_a, mu_b and tau_b in order the effective sample size becomes smaller, Rhat value is getting closer to 1 and showing poorer convergences. So tau_b shows significant issues during chain mixing and convergence.")

print("The traceplots indicate that mu_a and tau_a show good mixing with stable chains and reliable parameter estimates. Mu_b has moderate mixing with some fluctuations and higher variability, reflected in a lower effective sample size and a slightly higher Rhat value. In contrast, tau_b exhibits poor mixing with significant drifts and trends, indicating unreliable parameter estimates and convergence issues.")

print("Which model was estimated?")
print("The estimated model is most likely a hierarchical Bayesian regression model with random intercepts and slopes. It includes a Hierarchical structure with hyperparameters mu_a, tau_a, mu_b, and tau_b to model the population-level mean and standard deviation of the intercepts and slopes.")
print("What might be the cause of the convergence problems?")
print("Poor Mixing of Chains:")
print("The traceplots for tau_b show poor mixing with significant drifts and trends, indicating that the chains are not exploring the parameter space well.")
print("The effective sample size (n_eff) for tau_b is very low (~44), suggesting that the sampler struggles to produce independent samples.")

print("High Rhat Value:")
print("The Rhat value for tau_b is significantly above 1 (~1.106), indicating that the chains have not converged to the same distribution. This is a strong sign of convergence issues.")

print("Identifiability Issues:")
print("tau_b (the standard deviation for the random slopes) might be poorly identified, leading to difficulties in sampling. The data may not provide enough information to reliably estimate this parameter, resulting in convergence problems.")

print("Complex Model Structure:")
print("The hierarchical nature of the model, with both random intercepts and slopes, increases the complexity of the parameter space. This complexity can make it challenging for the sampler to converge, especially if the data do not strongly inform all parameters.")

print("Potentially Restrictive Priors:")
print("The priors specified for tau_b and other parameters might be too restrictive or not well-aligned with the data. This misalignment can hinder the sampler's ability to explore the parameter space effectively.")
```
