---
title: "Assignment 4"
format: pdf
execute: 
  cache: true
  warning: false
author:
  - name: "Kerem Karagöz"
  - name: "Immanuel Klein"
editor: 
  markdown: 
    wrap: sentence
---

GitHub: <https://github.com/immanuel-klein/bayesian-assignments.git>

```{r packages}
#load packages here
library(dplyr)
library(tidyverse)
library(ggplot2)
library(tinytex)
library(rethinking)
library(rstan)
```

```{r data}
# load the data set 'heart.csv' here
heart <- read.csv("heart.csv")
```

\newpage

# Task Set 1

## Task 1.1

Run a Bayesian logistic regression model to estimate the risk of men and women to develop a coronary heart disease (TenYearCHD).
Provide a summary of the posterior distributions.\
What is the average probability of men and women to develop the disease?

```{r m1}
#| echo: true
#| eval: true
#| output: false

# write data list and model here
heart.chd.gender <- na.omit(heart[, c("male", "TenYearCHD")])

model.gender <- ulam(
  alist(
    TenYearCHD ~ dbinom(1, p),  
    logit(p) <- a + bm * male,  
    a ~ dnorm(0, 1.5),
    bm ~ dnorm(0, 0.5)          
  ), data = heart.chd.gender, chains = 4, cores = 4
)
```

```{r 1.1}
#write code here
precis(model.gender, depth = 2)

samples <- extract.samples(model.gender)
cat("Avg. probability of CHD for women:", 
    round(mean(inv_logit(samples$a)), 3))
cat("Avg. probability of CHD for men:",
    round(mean(inv_logit(samples$a + samples$bm)), 3))
```

# Task 1.2

For the model of `Task 1.1`, visualize the posterior distribution of gender-differences to assess the credibility of the gender difference.

```{r 1.2}
samples.df <- data.frame(
  Female = inv_logit(samples$a),
  Male = inv_logit(samples$a + samples$bm)) %>%
  pivot_longer(cols = c(Female, Male), 
               names_to = "Gender", values_to = "Probability")

ggplot(samples.df, aes(x = Probability, fill = Gender)) +
  geom_density(alpha = 0.6) +
  labs(title = "Posterior of CHD Probability by Gender",
       x = "Prob. of CHD",
       y = "Density")
```

\newpage

# Task Set 2

## Task 2.1

Run a Bayesian logistic regression model to estimate the risk of men and women with and without diabetes to develop a coronary heart disease (TenYearCHD).
Provide a summary of the posterior distributions.
Does the effect of diabetes differ between men and women?

```{r m2}
#| echo: true
#| eval: true
#| output: false

# write data list and model here
heart.chd.gender.diabetes <- na.omit(
  heart[, c("male", "diabetes", "TenYearCHD")])

model.diabetes <- ulam(
  alist(
    TenYearCHD ~ dbinom(1, p),
    logit(p) <- a + bm * male + bd * diabetes + bmd * male * diabetes,
    a ~ dnorm(0, 1.5),
    bm ~ dnorm(0, 0.5),
    bd ~ dnorm(0, 0.5),
    bmd ~ dnorm(0, 0.5)
  ), data = heart.chd.gender.diabetes, chains = 4, cores = 4
)
```

```{r 2.1}
# write code here
# Summarize the posterior distributions
precis(model.diabetes, depth = 2)

samples <- extract.samples(model.diabetes)

cat("Avg. prob. of CHD for women without diabetes:", 
    round(mean(inv_logit(samples$a)), 3))
cat("Avg. prob. of CHD for men without diabetes:", 
    round(mean(inv_logit(samples$a + samples$bm)), 3))
cat("Avg. probability of CHD for women with diabetes:", 
    round(mean(inv_logit(samples$a + samples$bd)), 3))
cat("Avg. probability of CHD for men with diabetes:", 
    round(mean(inv_logit(samples$a + samples$bm + samples$bd + samples$bmd)), 
          3))
```

In both cases (diabetes vs. no diabetes) the probability of developing CHD is clearly higher for men than it is for women.

# Task 2.2

For the model of `Task 2.1`, visualize the posterior distributions of each group in one plot to better assess the credibility of the group differences.

```{r 2.2}
# write code here
samples.df <- data.frame(
  No.Diabetes.Female = inv_logit(samples$a),
  No.Diabetes.Male = inv_logit(samples$a + samples$bm),
  Diabetes.Female = inv_logit(samples$a + samples$bd),
  Diabetes.Male = inv_logit(samples$a + samples$bm + samples$bd + samples$bmd)
) %>%
  pivot_longer(cols = everything(), 
               names_to = "Group", 
               values_to = "Probability")

ggplot(samples.df, aes(x = Probability, fill = Group)) +
  geom_density(alpha = 0.6) +
  labs(title = "Posterior of CHD Probability by Group",
       x = "Probability of CHD",
       y = "Density")
```

\newpage

# Task Set 3

## Task 3.1

Run a Bayesian logistic regression model to estimate the effect of age on the risk of developing a coronary heart disease (TenYearCHD), separately for women and men.
Ensure that the regression intercept represents the risk of women and men with average age.
Provide a summary of the posterior distributions.

```{r m3}
#| echo: true
#| eval: true
#| output: false

# write data list and model here
heart.age <- na.omit(heart[, c("male", "age", "TenYearCHD")])

# Split data by gender and mean center. 
# We assume that the data needs to be centered for men and women seperately.
# _ notation instead of . for age_centered, 
# because ulam didn't accept age.centered.
heart.male <- heart.age %>% filter(male == 1) %>%
  mutate(age_centered = age - mean(age))
heart.female <- heart.age %>% filter(male == 0)%>%
  mutate(age_centered = age - mean(age))

model.male <- ulam(
  alist(
    TenYearCHD ~ dbinom(1, p),
    logit(p) <- a + bage * age_centered,
    a ~ dnorm(0, 1.5),
    bage ~ dnorm(0, 0.5)
  ), data = heart.male, chains = 4, cores = 4
)

model.female <- ulam(
  alist(
    TenYearCHD ~ dbinom(1, p),
    logit(p) <- a + bage * age_centered,
    a ~ dnorm(0, 1.5),
    bage ~ dnorm(0, 0.5)
  ), data = heart.female, chains = 4, cores = 4
)
```

```{r 3.1}
# write code here
precis(model.male, depth = 2)
precis(model.female, depth = 2)

samples.male <- extract.samples(model.male)
samples.female <- extract.samples(model.female)
```

## Task 3.2

For the model of `Task 3.1`, visualize the posterior distribution of differences in the age effect between women and men.
Does age increase the risk of developing the disease and does this effect differ between women and men?

```{r 3.2}
# write code here
samples.df <- data.frame(
  Male = samples.male$bage,
  Female = samples.female$bage
) %>%
  pivot_longer(cols = c(Male, Female), 
               names_to = "Gender", 
               values_to = "Age.Effect")

ggplot(samples.df, aes(x = Age.Effect, fill = Gender)) +
  geom_density(alpha = 0.6) +
  labs(title = "Posterior Distribution of Age Effect on CHD Risk by Gender",
       x = "Effect of Age on CHD Risk",
       y = "Density")

cat("Average age effect for men:", mean(samples.male$bage))
cat("Average age effect for women:", mean(samples.female$bage))
```

Age in fact increases the risk of developing CHD.
The influence that age has, is higher for women than it is for men.
